# Aesthetic mapping

::: {.callout-note}
## Class Objectives
- Understand the concept of aesthetic mapping
- Learn how to map variables to aesthetics in `ggplot2`
:::

## What is Aesthetic Mapping?
`aes()` stanfs for aesthetic mapping. It is a function in `ggplot2` that maps variables in your **data** to **aesthetics**. It tells the computer how to interpret the data to your plot. Anything that goes into the `aes()` function should be a column (a.ka. variable) in your data.

To see this, let us consider plotting the `body_mass_g` against the `bill_depth_mm` of the penguins dataset with all points colored by blue.

Let us compare to the following code:
::: {.panel-tabset}
## Mapping blue as aesthetic
```{r}
#| warning: false
library(ggplot2)
library(palmerpenguins)
ggplot(
    data = penguins,
    aes(
        x = body_mass_g,
        y = bill_depth_mm,
        color = "blue"
    )
) +
    geom_point() +
    theme_minimal()
```


```{r}
#| warning: false
library(ggplot2)
library(palmerpenguins)
ggplot(
    data = penguins,
    aes(
        x = body_mass_g,
        y = bill_depth_mm
    ),
    color = "blue"
) +
    geom_point() +
    theme_minimal()
```