# First Publication-Ready Figure

Instead of many ggplot2 tutorials that go direcly to other plot types, I instead decide to show you how to turn the plot you have into a publication-ready figure. We will use a variety of `ggplot2` family packages.

Let us first see what we left off in the previous chapter:
```{r}
#| warning: false
library(ggplot2)
library(palmerpenguins)
p <- ggplot(
    data = penguins,
    aes(
        x = bill_length_mm,
        y = bill_depth_mm,
        color = species,
        group = species
    )
) +
    geom_point() +
    geom_smooth() +
    theme_minimal()
p
```

# General Tips

::: {.panel-tabset}
## Adding Labels 
First, the labels are not clear. So it is unclear what the x-axis and y-axis represent. We can add labels to the axes using the `labs()` function. 
```{r}
#| warning: false
p + 
  labs(
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  )
```

## Adding a Title
Second, what is the figure about? We can add a title to the plot also using the `labs()` function:
```{r}
#| warning: false
p + 
  labs(
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    title = "Bill Length and Depth of Penguins are Positively Correlated"
  )
```


### Exercise {-}

```{r}
pacman::p_load(datasauRus)
```

We see that 
```{r}
datasaurus_dozen |>
  ggplot(aes(x = x, y = y, colour = dataset)) +
  geom_point() +
  theme_minimal() +
  facet_wrap(vars(dataset)) +
  labs(color = "Dataset")
```
<!-- https://tellingstorieswithdata.com/05-graphs_tables_maps.html -->


## Specialized TIps for Scatter Plots