# Visualizing Association

The association between variables is one of the most important questions in data analysis. 


We have already seen some examples of association in the previous chapters. But given its importance, we will greatly expand on this topic.

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(tidylog)
theme_set(hrbrthemes::theme_ipsum())
```

## Trend

We will use the `economics` dataset from the `tidyverse` package. 

```{r}
economics |>
    filter(month(date) == 1) |>
    ggplot(aes(x = date, y = unemploy)) +
    geom_point(color = "white", fill = "#0072B2", shape = 21, size = 2)
```

::: {.panel-tabset}

## Add line

```{r}
economics |>
    filter(month(date) == 1) |>
    ggplot(aes(x = date, y = unemploy)) +
    geom_line(size = 0.5, color = "#0072B2") +
    geom_point(color = "white", fill = "#0072B2", shape = 21, size = 2)
```


```

### Show difference between groups

### Stream graph


## Embed more information in the plot

Humans are bad at visualizing 3D data. Thus, we need to be careful when visualizing the association between variables.


## Dimension reduction

The data we typically work with is high-dimensional. We need to reduce the dimension of the data to visualize it.


