# Writing Functions

::: {.callout-note}
## Class Objectives
1. Understand when and why to write functions
2. Learn the basic structure of R functions
3. Master writing vectorized functions
4. Learn best practices for function documentation
:::

## Why Write Functions?

Remember our discussion about vectorization and nested data? We've been using functions all along - `mean()` and `sum()`. But sometimes, you need a function that doesn't exist yet. That's when you write your own!

Here are three signs you need a function:

1. You've copied and pasted code more than twice
2. You find yourself changing the same numbers in multiple places
3. You want to make your code more readable and maintainable

Writing functions also helps you avoid mistakes. For example, if you need to change a number in multiple places, you may forget to change one of them (speaking from experience :-(

## Basic Function Structure

Let's start with a simple function:

```{r}
calculate_bmi <- function(weight, height) {
    bmi <- weight / (height^2)
    return(bmi)
}

calculate_bmi(weight = 70, height = 1.75)
```

Let's break down the components:

1. `calculate_bmi`: The function name (make it descriptive!)
2. `function(weight, height)`: The arguments your function needs
3. The code block between `{` and `}`: What your function does
4. `return(bmi)`: What your function gives back

### Exercise: Your First Function {-}

Write a function to convert Fahrenheit to Celsius.

::: {.panel-tabset}

## Exercise

```{webr}
fahrenheit_to_celsius <- function(fahrenheit) {
    # Formula: (°F - 32) × 5/9 = °C
    ________
}

# Test your function
fahrenheit_to_celsius(32)  # Should return 0
fahrenheit_to_celsius(212) # Should return 100
```

## Solution

```{webr}
fahrenheit_to_celsius <- function(fahrenheit) {
    celsius <- (fahrenheit - 32) * 5/9
    return(celsius)
}

fahrenheit_to_celsius(32)
fahrenheit_to_celsius(212)
```
:::

## Writing Vectorized Functions

Remember how we talked about the importance of vectorization? When writing your own functions, you want them to be vectorized too! The good news is that R makes this easy - if you use vectorized operations inside your function, your function will be vectorized automatically. For example, our function `calculate_bmi()` is vectorized automatically:
 
```{r}
weights <- c(70, 80, 90)
heights <- c(1.75, 1.80, 1.85)
calculate_bmi(weights, heights)
```

Sometimes it might be not obvious that a function is vectorized. The most simple solution is to test your function to a vector or a tibble. If it works, it's (most likely) vectorized!

### Exercise: Vectorized Function {-}

Write a function to calculate bill area (length × depth) that works with the penguins dataset.

::: {.panel-tabset}

## Exercise

```{webr}
library(tidyverse)
library(palmerpenguins)

calculate_bill_area <- function(length, depth) {
    ________
}

# Test with the penguins dataset
penguins |>
    mutate(bill_area = calculate_bill_area(
        bill_length_mm, 
        bill_depth_mm
    ))
```

## Solution

```{webr}
library(tidyverse)
library(palmerpenguins)

calculate_bill_area <- function(length, depth) {
    return(length * depth)
}

penguins |>
    mutate(bill_area = calculate_bill_area(
        bill_length_mm, 
        bill_depth_mm
    ))
```
:::

::: {.callout-tip collapse="true"}

## Function Documentation

Good functions need good documentation. In R, we use **roxygen2** style comments to document our functions:

```{r}
#' Calculate Body Mass Index (BMI)
#'
#' @param weight Weight in kilograms
#' @param height Height in meters
#' @return BMI value (kg/m^2)
#' @examples
#' calculate_bmi(70, 1.75)
calculate_bmi <- function(weight, height) {
    bmi <- weight / (height^2)
    return(bmi)
}
```

The documentation includes:
1. A brief description
2. Parameter descriptions (`@param`)
3. What the function returns (`@return`)
4. Example usage (`@examples`)

:::

## Best Practices

When writing functions, follow these guidelines:

1. **Names Matter**
   - Use verbs for function names: `calculate_bmi()`, not `bmi()`
   - Be consistent with your naming style
   - Make names descriptive but not too long

2. **Arguments Matter**
   - Put the most important arguments first
   - Use clear argument names
   - Provide default values when it makes sense

```{r}
# Good function design
calculate_bmi <- function(weight, height, units = "metric") {
    if (units == "imperial") {
        # Convert pounds to kg and inches to meters
        weight <- weight * 0.453592
        height <- height * 0.0254
    }
    return(weight / (height^2))
}

# Now works with both metric and imperial units
calculate_bmi(70, 1.75) # metric (default)
calculate_bmi(154, 69, units = "imperial") # imperial
```

3. **Error Checking**
   - Check input validity
   - Provide helpful error messages

```{r}
calculate_bmi <- function(weight, height, units = "metric") {
    # Input validation
    if (weight <= 0 || height <= 0) {
        stop("Weight and height must be positive numbers")
    }

    if (units == "imperial") {
        weight <- weight * 0.453592
        height <- height * 0.0254
    } else if (units != "metric") {
        stop('Units must be either "metric" or "imperial"')
    }

    return(weight / (height^2))
}
```

### Exercise: Complete Function {-}

Write a complete function to calculate penguin body condition (mass divided by flipper length) with proper documentation and error checking.

::: {.panel-tabset}

## Exercise

```{webr}
#' Calculate penguin body condition
#' 
#' @param mass Body mass in grams
#' @param flipper_length Flipper length in mm
#' @return Body condition index
#' @examples
#' calculate_condition(3000, 200)
calculate_condition <- function(mass, flipper_length) {
    # Add input validation
    
    # Calculate condition
    
    # Return result
}

# Test with the penguins dataset
penguins |>
    mutate(condition = calculate_condition(
        body_mass_g,
        flipper_length_mm
    ))
```

## Solution

```{webr}
#' Calculate penguin body condition
#' 
#' @param mass Body mass in grams
#' @param flipper_length Flipper length in mm
#' @return Body condition index
#' @examples
#' calculate_condition(3000, 200)
calculate_condition <- function(mass, flipper_length) {
    # Input validation
    if (any(mass <= 0, na.rm = TRUE)) {
        stop("Mass must be positive")
    }
    if (any(flipper_length <= 0, na.rm = TRUE)) {
        stop("Flipper length must be positive")
    }
    
    # Calculate and return condition
    return(mass / flipper_length)
}

# Test with the penguins dataset
penguins |>
    mutate(condition = calculate_condition(
        body_mass_g,
        flipper_length_mm
    ))
```
:::

## Summary

Writing functions is a crucial skill that builds on our understanding of vectorization and nested data:

1. Write functions when you find yourself repeating code
2. Make your functions vectorized when possible
3. Use clear names and good documentation
4. Include input validation and helpful error messages
5. Test your functions with simple cases first

Remember: A well-written function is like a good tool - it should do one thing, do it well, and be easy to use! 🛠️
