# Using map() and Friends

::: {.callout-note}
## Class Objectives
1. Understand the concept of mapping functions
2. Learn the basic `map()` functions and their variants
3. Master working with lists using `map()`
4. Learn when to use `map()` vs. `rowwise()`
:::

## What is Mapping?

Remember how we used `rowwise()` to apply functions to nested data? Well, `map()` is like its cool cousin - it's another way to apply functions to multiple pieces of data. Think of it as saying "take this function and apply it to each element in this list".

### The Basic Idea

Let's start with a simple example. Suppose you want to calculate the length of several words:

```{r}
#| message: false
library(tidyverse)

# The old way (one by one)
nchar("apple")
nchar("banana")
nchar("cherry")

# The map way
words <- c("apple", "banana", "cherry")
map(words, nchar)
```

See what happened? `map()` took our `nchar` function and applied it to each word in our list. Magic! 🪄

### Different Types of map()

The `map()` family has several variants, depending on what type of output you want:

```{r}
# map() returns a list
map(words, nchar)

# map_dbl() returns a numeric vector
map_dbl(words, nchar)

# map_chr() returns a character vector
map_chr(words, toupper)

# map_lgl() returns a logical vector
map_lgl(words, function(x) nchar(x) > 5)
```

Think of these variants as different containers for your results:
- `map()`: A flexible shopping bag (can hold anything)
- `map_dbl()`: A number-only box
- `map_chr()`: A text-only box
- `map_lgl()`: A true/false box

## Working with Lists

Let's get a bit more practical. Remember our penguin friends?

```{r}
library(palmerpenguins)

# Create a list of data for each species
species_data <- penguins |>
    group_by(species) |>
    nest() |>
    pull(data) # <1>
```
1. `pull(data)` extracts just the data column, giving us a list of tibbles

Now we can use `map()` to analyze each species:

```{r}
# Calculate mean bill length for each species
map_dbl(species_data, function(df) mean(df$bill_length_mm, na.rm = TRUE))
```

### Using the Formula Shorthand

Writing anonymous functions can be tedious. Luckily, `map()` has a neat shortcut:

```{r}
# The long way
map_dbl(species_data, function(df) mean(df$bill_length_mm, na.rm = TRUE))

# The short way
map_dbl(species_data, \(df) mean(df$bill_length_mm, na.rm = TRUE))

# The even shorter way (using formula notation)
map_dbl(species_data, ~ mean(.x$bill_length_mm, na.rm = TRUE))
```

All three do exactly the same thing! The formula notation (with `~` and `.x`) is a purrr special sauce that makes your code more concise.

### Exercise: Species Statistics {-}

Calculate the standard deviation of bill depth for each species.

::: {.panel-tabset}

## Exercise

```{webr}
library(tidyverse)
library(palmerpenguins)

species_data <- penguins |>
    group_by(species) |>
    nest() |>
    pull(data)

# Calculate sd of bill depth
map_dbl(species_data, ~ ________)
```

## Solution

```{webr}
library(tidyverse)
library(palmerpenguins)

species_data <- penguins |>
    group_by(species) |>
    nest() |>
    pull(data)

map_dbl(species_data, ~ sd(.x$bill_depth_mm, na.rm = TRUE))
```
:::

## Complex Operations with map()

Let's fit a linear model for each species, just like we did with `rowwise()`:

```{r}
# Fit models
models <- map(species_data, ~ lm(bill_length_mm ~ bill_depth_mm, data = .x))

# Extract R-squared values
map_dbl(models, ~ summary(.x)$r.squared)
```

Notice how we:
1. Used `map()` to create a list of models
2. Used `map_dbl()` to extract numeric R-squared values
3. Didn't need to worry about list columns or `rowwise()`

### Multiple Inputs with map2()

Sometimes you need to work with two lists at once. That's where `map2()` comes in:

```{r}
x <- list(1, 2, 3)
y <- list(10, 20, 30)

map2(x, y, ~ .x + .y)  # .x is from first list, .y from second
```

### Exercise: Custom Models {-}

Fit a linear model for each species using different formulas.

::: {.panel-tabset}

## Exercise

```{webr}
library(tidyverse)
library(palmerpenguins)

# Data
species_data <- penguins |>
    group_by(species) |>
    nest() |>
    pull(data)

# Formulas for each species
formulas <- list(
    bill_length_mm ~ bill_depth_mm,
    bill_length_mm ~ flipper_length_mm,
    bill_length_mm ~ body_mass_g
)

# Fit different models for each species
map2(species_data, formulas, ~ ________)
```

## Solution

```{webr}
library(tidyverse)
library(palmerpenguins)

# Data
species_data <- penguins |>
    group_by(species) |>
    nest() |>
    pull(data)

# Formulas for each species
formulas <- list(
    bill_length_mm ~ bill_depth_mm,
    bill_length_mm ~ flipper_length_mm,
    bill_length_mm ~ body_mass_g
)

# Fit different models for each species
map2(species_data, formulas, ~ lm(.y, data = .x))
```
:::

## map() vs. rowwise(): When to Use Which?

Both `map()` and `rowwise()` can handle similar tasks, but they have their sweet spots:

Use `map()` when:
1. You're working with plain lists
2. You want to chain operations with the pipe
3. You need different output types (numeric, character, etc.)
4. You want to use the formula shorthand

Use `rowwise()` when:
1. Your data is already in a tibble
2. You need to maintain the tibble structure
3. You're working with multiple columns
4. You want to keep the results alongside the original data

## Summary

The `map()` family of functions gives you a powerful way to work with lists and apply functions to multiple elements. Remember:

1. Use the right map variant for your output type:
   - `map()` for lists
   - `map_dbl()` for numbers
   - `map_chr()` for text
   - `map_lgl()` for true/false

2. Take advantage of the formula shorthand (`~` and `.x`) for cleaner code

3. Use `map2()` when you need to work with two lists at once

4. Choose between `map()` and `rowwise()` based on your data structure and needs

In the next chapter, we'll explore even more powerful functions from the purrr package! 🚀
