# Making R Faster 🚀


::: {.callout-note}
## Class Objectives
1. Measure your code’s speed – Know exactly how fast (or slow) your code is running
2. Parallel computing magic – Get free speed boosts by using multiple CPU cores
3. Turbocharge your data wrangling – Swap tidyverse functions with faster engines under the hood.
:::

R sometimes gets a bad rap for being slower than languages like C++ or Julia. That's partly true: if you are really serious about performance, you might want to learn another language. However, **you can make R much faster without rewriting your code or learning a new language.** 

Let's start by loading tidyverse and a dataset from the `nycflights13` package which has more than 330,000 rows.

```{r}
#| message: false
#| warning: false

library(tidyverse)
data(flights, package = "nycflights13")
```

Note that we are not using the `tidylog` package this time. `tidylog` marks up the verbs with comments so that you know what happens in each step. This is great for understanding what the code is doing, but it slows down the execution. 

## Figure Out How Fast Your Code Is 

### Quick Check with `tictoc`

The `tictoc` package (no relation to social media platform) lets you time code chunks easily:

```{r}
#| message: false
#| warning: false

library(tictoc)

tic() # <1>

model_r2 <- flights |> # <2>
  group_by(carrier, origin, year) |> # <2>
  nest() |>   # <2>
  mutate(r2 = map_vec(data, \(x) summary(lm(arr_delay ~ dep_delay, data = x))$r.squared)) # <2>

toc() # <3>
```
1. Start the timer!
2. Your code goes in between `tic()` and `toc()`.
3. Stop the timer and print the time

### Compare Methods with `microbenchmark`

Want to test multiple approaches? `microbenchmark` runs code repeatedly and compares results:

```{r}
#| warning: false
library(microbenchmark)

benchmark <- microbenchmark(
  method_rowwise = flights |>  # <1>
    group_by(carrier, origin, year) |> 
    nest() |> 
    rowwise() |>
    mutate(r2 = summary(lm(arr_delay ~ dep_delay, data = data))$r.squared) |>
    ungroup(),
  method_map = flights |>  # <2>
    group_by(carrier, origin, year) |> 
    nest() |>   
    mutate(r2 = map_vec(data, \(x) summary(lm(arr_delay ~ dep_delay, data = x))$r.squared))
  ,
  times = 10 # <3>
) 

print(benchmark) # <4>
```
1. Name the method on the left and the code on the right.
2. Use another method
3. Run each code 10 times
4. Print the time it took to run each method

It might be helpful to visualize the results. You can use the `autoplot` function to do this easily.

```{r}
#| warning: false

autoplot(benchmark)
```

### Estimate how long it will take to run your code

A great new feature in `purrr` is the introduction of `.progress` argument. This allows you to estimate how long it will take to run your code.

As the codes above runs very fast, we use a more time-consuming task (just for illustration purposes).
```{r}
#| message: false
#| warning: false
#| eval: false

tibble(x = 1:500) |>
  mutate(y = map_dbl(x, \(x) {
    Sys.sleep(0.01) # <1>
    x
  },
  .progress = TRUE)) # <2>
```
1. Sleep for 0.01 seconds
2. Show a progress bar

## Parallel Computing

R usually uses one CPU core. But most computers have 4-8 cores! By sending the same task to multiple cores, you get automatic speedups for free! 

We can use the `furrr` package to parallelize the map functions. When setting the number of workers, be sure to leave one core for the leader (the one that coordinates the workers).

```{r}
#| message: false
#| warning: false

library(furrr)
parallelly::availableCores() # <1>
plan(multisession, workers = 4)  # <2>
tic()
model_r2 <- flights |> 
  group_by(carrier, origin, year) |> 
  nest() |> 
  mutate(r2 = future_map_dbl(data, \(x) summary(lm(arr_delay ~ dep_delay, data = x))$r.squared))
toc()
```
1. Check how many cores are available
2. Set the number of workers to 4

The syntax is almost identical to `purrr`, just replace `map` with `future_map` (you can even use the `.progress` argument to show a progress bar), and set before the code how many cores you want to use (set it lower than the number of cores on your computer).

Note: Parallelizing adds some overhead (think of the leader needs time to coordinate the workers). Thus, it may cost more time than the non-parallelized version for small datasets or quick calculations.

::: {.callout-tip collapse="true"}

## The magic of `mirai`


This is a feature in development version of `purrr` ([version 1.0.4.0](https://purrr.tidyverse.org/dev/reference/parallelization.html)). However, I have little doubt that it will be merged into the main version anytime soon: it is so much easier to use with and is up to 1000 times faster compared to `furrr`.

To try it out, you need to install the development versions of `purrr` and `mirai`:

```{r}
#| eval: false
pak::pak("tidyverse/purrr")
pak::pak("shikokuchuo/mirai")
```

Then we simply need to add `.parallel = TRUE` to the `map` function.

```{r}
#| message: false
#| warning: false

library(purrr)
library(mirai)
mirai::daemons(4) # <1>

tic()

model_r2 <- flights |> 
  group_by(carrier, origin, year) |> 
  nest() |> 
  mutate(r2 = map(data, 
                  \(x) x$arr_delay - x$dep_delay,
                  .parallel = TRUE)) # <2>

toc()
```
1. Set the number of cores you want to use
2. Add `.parallel = TRUE` to the `map` function

:::

## Swap Tidyverse for Speed Demons

::: {.panel-tabset}

## `dtplyr` (Tidy Grammar + `data.table` Speed)

`data.table` is blazing fast but has a steep learning curve. `dtplyr` lets you write tidyverse code that gets translated to `data.table` behind the scenes:

```{r}
#| message: false
#| warning: false
#| eval: false
library(dtplyr)

flights |> 
  lazy_dt() |>  # <1>
  group_by(carrier, origin, year) |> # <2>
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) |> # <2>
  as_tibble() # <3>
```
1. Convert the tibble to a data table
2. Use all the `dplyr` functions you know
3. Convert the data table back to a tibble

## `duckplyr` (Tidy Grammar + DuckDB Speed)

DuckDB is a fast database engine, which works super well if you have a lot of data (like millions of rows). `duckplyr` is a package that translates `dplyr` code to `duckdb` code behind the scenes. So again, you can write as usual and do not need to worry about the underlying heavy machinery.

The messages when you load the package clealy tell you that the `dplyr` functions are overridden.
```{r}
library(conflicted) # <1>
library(duckplyr) # <2>
```
1. Resolve conflicts
2. Load the package and override the `dplyr` functions

Then we just do what we always do.
```{r}
#| eval: false
flights |>
  group_by(carrier, origin, year) |> 
  summarize(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) 
```


:::

## Additional Resources

Making code faster is a very broad topic. Here are some additional resources that you need even more speed:

- [collapse](https://sebkrantz.github.io/collapse/)
- [tidypolars](https://www.tidypolars.etiennebacher.com/)
- [arrow](https://arrow.apache.org/docs/r/index.html)

All of these packages are under active development, and it is hard to say which one is the best (they all have their own strengths). The good news is that they all follow the tidyverse grammar (so very mild learning curve).

## Remember: Speed Isn’t Everything! 🛑

> Premature optimization is the root of all evil.
> – Donald Knuth

Only optimize code that’s slow enough to annoy you. Focus on correctness and readability first!




