# Put into practice

How many times have you heard that "success is because of hard work"? It would be fun to consider a world where success is completely random: If everyone starts equally, and all successes (like wealth) are completely random, what will happen? 

In this world, will everyone end up with roughly the same amount of wealth? Or do we still see a sharp distribution of wealth?

Let's find out!

```{r}
#| message: false
#| warning: false
library(tidyverse)

num_players <- 45
rounds <- 5000
init_wealth <- 45

money <- tibble(
    time = 0,
    person = 1:num_players,
    wealth = init_wealth
)

# initialize the bank
# columns wealths of the NUMPLAYERS players
# rows show wealths of each of the ROUNDS ticks of the clocks
bank <- matrix(0, nrow = ROUNDS, ncol = NUMPLAYERS)
bank[1, ] <- c(rep(INITWEALTH, NUMPLAYERS))

# function to give a dollar to someone other than oneself
get_recipient <- function(player) {
    sample(setdiff(1:NUMPLAYERS, player), 1)
}

# execute trades and update the ledger
for (i in 2:ROUNDS) {
    # every player with wealth chooses another person to receive a buck
    recipients <- sapply(which(bank[i - 1, ] > 0), get_recipient)

    # table of the dollars owed each person
    count_table <- table(recipients)

    # get the indices of the people owed money
    indices <- as.integer(names(count_table))

    # everyone gives up a dollar, unless they are at zero
    bank[i, ] <- ifelse(bank[i - 1, ] > 0, bank[i - 1, ] - 1, bank[i - 1, ])

    # selected people receive dollars
    bank[i, indices] <- bank[i, indices] + count_table
}

#################### Animate it
# Make a suitable long data frame
df <- as.data.frame(bank)
names(df) <- 1:NUMPLAYERS
df <- df %>%
    mutate(frame = 1:ROUNDS) %>%
    gather(person, wealth, 1:NUMPLAYERS) %>%
    mutate(person = as.numeric(person)) %>%
    arrange(frame) %>%
    group_by(frame) %>%
    mutate(rank = rank(wealth, ties.method = "random")) %>%
    ungroup() %>%
    gather(histtype, playerid, c(person, rank)) %>%
    mutate(histtype = sprintf("Ordered by %s", histtype))
```


```{r}

df |>
    filter(frame == 5000 & histtype == "Ordered by person") |>
    ggplot(aes(x = reorder(playerid, wealth), y = wealth)) +
    geom_bar(stat = "identity")
```

